--- 
title: 'TMA4315: Compulsory exercise 2 Logistic regression and Poisson regression'
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
date: "`r format(Sys.time(), '%d.%m.%Y')`"
subtitle: 'Group XX: Henrik Syversveen Lie, Mikal Stapnes, Oliver Byhring'
---


```{r setup, include = FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)
```

# Part 1: Logistic regression

## a)

We let $y_i$ be the number of successfull ascents, and $n_i$ be the total number of attempts (success + fail) of the i'th mountain. We then do binary regression with the logit link to model the probability of success. This gives

1. Model for response: $Y_i \sim \text{Bin}(n_i,\pi_i), \quad \text{for } i = 1,\dots,113$
2. Linear predictor: $\eta_i = {\bf x} _i^T \beta$
3. Link function: $\eta_i = \ln(\frac{\pi_i}{1-\pi_i})$

where $x_i$ is a $p$ dimensional column vector of covariates for observation $i$, and $\beta$ is the vector of regression parameters.

# Part 2: Poisson regression - Eliteserien 2018

## a)
```{r, echo = F, eval = T}
filepath <- "https://www.math.ntnu.no/emner/TMA4315/2018h/eliteserien2018"
eliteserie <- read.table(file = filepath, header = TRUE, colClasses = c("character", 
    "character", "numeric", "numeric"))
head(eliteserie)
dim(eliteserie)
contingency = matrix(rep(0,25),nrow = 5, ncol = 5)
rownames(contingency) = c("0","1","2","3","4+")
colnames(contingency) = c("0","1","2","3","4+")
print(contingency[eliteserie[2,3],eliteserie[2,4]])

for (i in 1:192){
  if (eliteserie[i,3]<4 & eliteserie[i,4]<4){
    contingency[eliteserie[i,3]+1,eliteserie[i,4]+1] = contingency[eliteserie[i,3]+1,eliteserie[i,4]+1]+1
  }  else if (eliteserie[i,3]<4 & !eliteserie[i,4]<4){
    contingency[eliteserie[i,3]+1,5] = contingency[eliteserie[i,3]+1,5] + 1
  }  else if (!eliteserie[i,3]<4 & eliteserie[i,4]<4){
    contingency[5,eliteserie[i,4]+1] = contingency[5, eliteserie[i,4]+1] + 1
  }  else{
    contingency[5,5] = contingency[5,5] + 1
  }
}


```











